---
title: 'The effect of aging on visuomotor processes: Analyses and Results'
author:
- Chad Vachon
- Shanaathanan Modchalingam
- Marius 't Hart
- Denise Henriques
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: '2'
---

# Preliminaries

```{r eval=T, echo=F, message=F}

# these libraries are used later on

library('ez')        # ANOVAs
library('car')       # ANOVAs
library('afex')      # ANOVAs
library('emmeans')   # post-hoc contrasts for ANOVAs
library('svglite')   # SVG figures

# if they can't be loaded, install them, by uncommenting this:
# install.packages(c('ez','car','afex','emmeans','svglite'))

```

Functions for R scripts that we need to use:

```{r}
source('R/learning_curves.R') # this file has all the data analysis stuff for learning curves
source('R/reach_aftereffects.R') # in this file we deal with reach aftereffects
source('R/localization.R') # scripts for localization stuff
source('R/shared.R') # color scheme, function for confidence intervals, data downloader
```

Check to see if data is present and if not, populate folder:

```{r}
ensureData(check=TRUE)
```

We'll look at some participant descriptives first:

```{r demographics}
participants <- checkParticipantsData()
# other analyses could be done with these demographics
groups <- aggregate(curves ~ group, data=participants, FUN=sum)
names(groups)[which(names(groups) == 'curves')] <- 'N'
groups$age_mean <- aggregate(age ~ group, data=participants, FUN=mean)$age
groups$age_std <- aggregate(age ~ group, data=participants, FUN=sd)$age
print(groups)
```

As well as percentages of rejected trials:

```{r}
rejected <- rejectedData()
print(rejected$full)
```

That's too much to report, so here is a summary table:

```{r}
print(rejected$short)
```

Slightly more trials have been rejected for the older adults as for the younger adults, but all percentages are pretty low (well below 5%).

# Learning

In this section, we analyze how older and younger participants' adapt their hand movements with or without instruction.  
Instructions should affect early learning for both age groups, but do they show whether instructions help older adults as much as younger adult? 

## Was there any learning?

We want to know whether or not reach direction changed during the rotated training block for all groups and whether this varied with instruction type and age.

The following is a plot of the reaching data in the first 90 trials ("learning curves"):

```{r, fig.width=8, fig.height=5}
# produce a plot (current Figure 3)
plotLearningCurves()
```

Reach deviations of 30 degrees would fully counter the rotation, so it looks like all groups did equally well by the end of training, that instruction helped in the first few trials, but also that older adults perhaps benefit less from the instructions.

We do a three-way mixed ANOVA on reach deviations, using block (first, second and last), instructed (non-instructed or instructed) and agegroup (younger or older) as factors.

```{r}
# do a statistical analysis (omnibus ANOVA)
learningCurveANOVA()

```

We find a significant effect of block (that separately interacts with instructions and with age group) so that reach deviations changed throughout the rotated training. We also show that instruction type does interact with age group, so that instructed older and younger participants perhaps learned differently in the training.

## First Block of Training

We had a specific prediction that instructed older adults would not benefit from the instructions as much as instructed younger adults. Here we want to see the effects of and interaction between age group and instruction, very early on in learning.


```{r}
blockLearningANOVA(block=1)
```

There is an interaction between age group and instruction, so we want to test what causes this. Hence we do some follow-up tests. First a t-test looking to see if instruction had an effect within older adults.


```{r}
blockLearningTtest(block=1, 
                   groups=list(
                     list('agegroup'='older', 'instructed'=TRUE),
                     list('agegroup'='older', 'instructed'=FALSE)))
```

It appears, an instruction **did** benefit older adults. Therefore, we now want to see if it benefited older adults as much as younger adults:

```{r}
blockLearningTtest(block=1, 
                   groups=list(
                     list('agegroup'='older', 'instructed'=TRUE),
                     list('agegroup'='younger', 'instructed'=TRUE)))
```

There is a difference in that older adults do not benefit from instruction as much as younger adults - at least in the first few trials.

## Second Block of Training

We wanted to see whether the effect of age and instruction would continue for the second block of trials. We repeat the ANOVA on reach deviations using age group and instruction as factors, but now on the data from the second block of three trials:

```{r}
blockLearningANOVA(block=2)
```

Crucially, there is no effect of age group, nor an interaction between age group and instruction. Hence we skip any follow-ups: the benefit of instruction is not different for the two age groups in the second block of three trials.

# Reach Aftereffects

After training with a visual rotation, with or without instructions, people are more or less aware of the rotation that the experiment imposes on the visual feedback. To the degree that they are aware of what they are doing to reach to the target with altered feedback, they should also be able to "switch off" their strategy, which we test here with a PDP-like method (see Werner et al., 2015) by asking participants to reach with or without a strategy in the absence of visual feedback ("no-cursor" or "open-loop" reaches).


```{r, fig.width=6, fig.height=5.5}
# produce a plot (current Figure 4)
plotReachAftereffects()
```

## Are there reach aftereffects?

First, we use the "without strategy" condition to test if there are effects of rotated training on no-cursor reach directions, in a three-way mixed ANOVA, using training session (aligned or rotated-with-strategy) as a within-subjects factor, and instruction (yes or no) and age group (older or younger) as a between-subjects factor:

```{r}
NoCursorANOVA()
```

As expected there is a main effect of training, i.e. there were reach aftereffects. In other words: the participants adapted to the rotation. There were other effects which we will explore below, but first we subtract aligned reach deviations from those in the rotated session.

It is also important to notice that while instruction interacted with age-group and with both age-group and session, it did not have a main effect or an interaction with instruction alone. This suggests that explicit learning, as induced by instruction did not decrease (implicit) reach aftereffects. To test this we ran two single-sided t-tests checking if reach aftereffects were smaller for the instructed as compared to the non-instructed participants for both age groups:

```{r}
NoCursorTtests()
```

The lack of an effect suggests that explicit learning by instruction does not decrease reach aftereffects in either age group. Looking at the figure, it seems that instructed older adults even have slightly larger reach aftereffects than non-instructed older adults.

## Are reach aftereffects affected by strategy use in the four groups?

We test if the magnitude of reach aftereffects is different given age group or instruction, by doing a three-way, mixed design ANOVA on reach aftereffects with instructed (yes or no) and age group (older vs. younger) as between-subjects factors and strategy use (with or without) as a within-subjects factor:

```{r}
# do a statistical analysis (ANOVA on reach aftereffects)
RAE.ANOVA()
```

Instruction and strategy have an effect and interact. When we look at the figure showing the magnitude of reach afterreffects, we can see that this means that when provided with an instruction, people can elicit part of their learning at will, i.e.: learning is (partially) made explicit by the instructions.

There is a main effect of age group, this doesn't interact with instruction type (or strategy), which would mean that the effect of instruction on explicit learning is the same for both age groups.

### Post-hoc: main effect of age

However, the main effect of age could mean that implicit learning has inrceased overall in the older participants. Looking at the figure this might be driven solely by the reach aftereffects of instructed older adults in the without-strategy condition. To test this, we run some post-hoc tests. First, a set of 6 t-tests in the without strategy condition comparing all groups:

```{r}
RAE.posthoc(doEmmeans=FALSE, doSubAOVs=FALSE, doTtests=TRUE)
```

All comparisons including the instructed older adults were significant, and none of the others. (If FDR correction for multiple comparisons is used, the comparison between older instructed and older non-instructed without-strategy reach aftereffects is no longer significant.) These results confirm that the main effect of age might be driven by one of the four older-adult data points, and not be generic.

Second, we can split the 2x2x2 ANOVA into four 2x2 ANOVAs, which each preserve a comparison between age groups. We can test in the exclude strategy condition, in the include strategy condition across all four groups, or we can test in the instructed groups or the non-instructed groups across both the include and exclude strategy conditions. If the instructed older adults' responses in the without strategy condition are driving the main effect of age, we should only see effects in the without strategy condition comparing all groups (perhaps interacting with instruction, as the non-instructed older adults don't have higher implicit reach aftereffects), as well as in the ANOVA on the two instructed groups.

```{r}
RAE.posthoc(doEmmeans=FALSE, doSubAOVs=TRUE, doTtests=FALSE)
```

These ANOVAs confirm that age-affects are only present when data from the instructed older adults in the without-strategy condition is included. While still other post-hoc tests are possible, these already show that the instructed older adults' without-strategy reach aftereffects are driving this effect. The question would now be if this is a real effect or an outlier, but we do not have the data to answer this.

# Localization

To move we need to have position information on our limbs. In the absence of vision we can use proprioception as well efference-based predicted consequences. Both of these are changed with visuomotor training. Here we try to tease them apart with our passive and active localization task.

## Does localization change with training?

Here we assess whether older and younger adults differ in how they estimate the location of their hand after training. We also determine whether this is different when they have access only to proprioception (passive localization) compared to when they have access to prediction and proprioception (active localization). The difference between these two is the predicted consequences plot below.

```{r, fig.width=8, fig.height=6}
# produce a plot: current Figure 5
plotLocalization()
```

Shown is the training-induced shift in localization for active (panel A) and passive localization (panel B). We show the data across the workspace, as well as averages across three discrete points in the workspace (50, 90 and 130 degrees), each with a 95% confidence interval. In panel C the difference between active and passive localization shifts is shown. The idea is that if this difference deviates from zero (active and passive localization are systematically different), then predicted sensory consequences contributed to active localization.

We can see that all groups show some kind of shift in both active and passive localization, with perhaps larger shifts for the older adults (we'll test this below). The third panel seems to show that younger adults may rely somewhat on predicted sensory consequences while older participants didn't, but this needs testing (see below).

First we want to see if there is any effect of rotated training on localization, so we do an mixed-design "omnibus ANOVA" using rotated (no vs. yes), passive movement (no=active vs. yes=passive) as within-subject factors and age group (older vs. younger) and instruction (yes or no) as between-subject factors.

```{r}
# do a statistical analysis (omnibus ANOVA)
localizationANOVA()
```

Rotated training shifts localization. There are some other effects, that we will test in more depth later on. Either way, we now calculate localization shifts as the difference between aligned and rotated localization responses.

If explicit learning prevents updating predicted sensory consequences or recalibrating proprioception (since you **know** the feedback is not veridical) then instructions should affect localization shift. Also, if older participants use instructions differently, or if their proprioceptive acuity has decreased, age group should also affect localization shift. That is what we test here: 

```{r}
# do a statistical analysis (ANOVA on passive localization difference scores)
localizationANOVA(test='shifts')
```

There is a main effect of age as well as a main effect of movement type, but no interaction. That is, it seems that our observation that older adults had larger training-induced shifts of localization is true. The effect of passive versus active localization seems to suggest that (at least for some groups) active localization was influenced by predicted sensory consequences.

It is interesting to see that instruction has no effect. This means that any effect of instruction on hand localization is minimal, so it seems that the signals contributing to hand localization are primarily implicit.

## Is proprioceptive recalibration affected by age and instruction?

We saw an effect of age group above. Here we test if this effect persists in passive localization alone, and also verify that instruction had no effect.

```{r}
localizationANOVA(test='passive')
```

There is an effect of age group but no other effect. This confirms that older adults show larger proprioceptive recalibration.

## Are predicted sensory consequences affected by age and instruction?

Due to experience, older adults may be able to use the predicted consequences of a motor command to better estimate where their hand is. Here we assess if age does indeed lead to this effect.

We can see in our plot that older adults predicted sensory consequences overlap with the zero on the axes, wheras younger adults have predicted consequences that are slightly deviated from zero.

We run an ANOVA with age and instruction as factors to determine whether these factors play a role in our hand location estimates. 

```{r}
# do a statistical analysis (ANOVA on the difference between active and passive localization difference scores)
predictedConsequencesANOVA()

```

We find no effect of any of these factors. 

## Do older adults show predicted sensory consequences at all?

Given the above results, we wanted to see if there is any evidence that older adults update predicted sensory consequences at all.

We test the magnitude of predicted sensory consequences against 0 for all younger as well as for all older participants with t-tests.


```{r}
predConsTtests()
```
Here we find that they do not, while younger adults do (effect size is reasonable in the younger adults, but power is low in each case). We can only speculate why this is the case.

# Published figures

The chunk of code below generates the svg versions of the figures that we submitted to the journal:

```{r createSVGs, include=TRUE, results="hide"}

plotLearningCurves(target='svg')

plotReachAftereffects(target='svg')

plotLocalization(target='svg')

```

In order to reproduce the analyses, it might be useful to have version information for all packages used:

```{r}
sessionInfo()
```

